Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer

### 2020-03-04 manuscript/manuscript.tex
included more of Fabienne's and Nadine's comments

\documentclass[twocolumn, 10pt]{article}
\usepackage{amsmath,amssymb}
\usepackage[format=hang,labelfont=bf,textfont=small,singlelinecheck=false,justification=raggedright,margin={20pt,20pt},figurename=Fig.]{caption}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage{threeparttable}
\usepackage[numbers]{natbib}
\newcommand{\mean}[1]{$\overline{\mbox{#1}}$}
\newcommand{\median}[1]{$\hat{\mbox{#1}}$}

\makeatletter
\def\affiliation#1{\gdef\@affiliation{#1}}
\def\abstract#1{\gdef\@abstract{#1}}
\def\graphabst#1{\gdef\@graphabst{#1}}
\def\keywords#1{\gdef\@keywords{#1}}
\def\corresp#1{\gdef\@corresp{#1}}
\def\bioauthor#1#2{\centerline{\textbf{#1}}\par #2}
\newcommand{\MakeTitle}{
  \newpage
  \begin{center}
    \Large \@title\par
    \large \@author
  \end{center}
  \@affiliation\par
  \@abstract\par
  \vskip 1em
}
\makeatother

\title{Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer}
\author{Héctor Climente-González$^{1,2,3}$, Christine Lonjou$^{1,2,3}$, Fabienne Lesueur$^{1,2,3}$,\\
GENESIS Study collaborators, Dominique Stoppa-Lyonnet$^{4,5,6}$,\\
Nadine Andrieu$^{1,2,3}$, Chloé-Agathe Azencott$^{3,1,2}$}
\date{$^{1}$Institut Curie, PSL Research University, F-75005 Paris, France;\\
  $^{2}$INSERM, U900, F-75005 Paris, France;\\
  $^{3}$MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France;\\
  $^{4}$Service de Génétique, Institut Curie, F-75005 Paris, France;\\
  $^{5}$INSERM, U830, F-75005 Paris, France;\\
  $^{6}$Université Paris Descartes.\\
}
\hypersetup{
  pdfauthor={Héctor Climente-González, Christine Lonjou, Fabienne Lesueur, GENESIS Study collaborators, Dominique Stoppa-Lyonnet, Nadine Andrieu, Chloé-Agathe Azencott},
  pdftitle={Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer},
  pdfkeywords={},
  pdfsubject={},
  pdflang={English}}
\begin{document}

@@

\Large\textbf{Abstract}

Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. It does so by jointly considering the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study the utility of six network methods to discover new biomarkers for breast cancer susceptibility by searching subnetworks highly associated to this phenotype. We interrogate a familial breast cancer genome-wide association study (GWAS) focused on \emph{BRCA1/2} negative French women. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. Interestingly, a combination of solution subnetworks provided a concise subnetwork of 93 genes, enriched in known breast cancer susceptibility genes (\emph{BABAM1}, \emph{BLM}, \emph{CASP8}, \emph{FGFR2}, and \emph{TOX3}, Fisher's exact test P-value = $7.8 \times 10^{-5}$ ) and more central than average. Additionally, it includes subnetworks of mechanisms related to cancer, like protein folding (\emph{HSPA1A}, \emph{HSPA1B}, and \emph{HSPA1L}) or mitochondrial ribosomes (\emph{MRPS30}, \emph{MRPS31}, \emph{MRPS18B}). We also observed a general dysregulation in the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. By trading statistical astringency for biological meaningfulness, most network methods get more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility.

@@

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, the experimental setting also presents intrinsic challenges. Some of them stem from the high-dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with large and moderate effects can be detected. And it is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically Bonferroni. However, Bonferroni is known to be overly conservative when the statistical tests are correlated, as is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. On top of that, recent large-sampled studies suggest that most of the genome contributes to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The recently proposed omnigenic model \cite{boyle_expanded_2017} offers an explanation: genes are very functionally inter-related and influence each other's behavior, which allows alterations in most genes to impact the ``core'' genes directly involved in a disease's mechanism. Hence, a comprehensive statistical framework which includes the structure of biological data might address the aforementioned issues.

In this regard many authors turn to network biology to handle the complex interplay of biomolecules that lead to disease \cite{furlong_human_2013}. As its name suggests, network biology models biology as a network, where the biomolecules under study, often genes, are nodes, and between them are the edges that link them. These functional relationships come from evidence that the genes jointly contribute to a biological function; for instance, their expression is correlated, or their products establish a protein-protein interaction. Under this view, complex diseases are not the consequence of a single altered gene, but of the interaction of multiple interdependent molecules \cite{barabasi_network_2011}. In fact, an examination of biological networks shows that disease genes have differential properties \cite{barabasi_network_2011} \cite{pinero_uncovering_2016}. This is particularly true for cancer driver genes, which tend to be key players in connecting different, densely-connected communities of genes. Additionally, as genes that contribute to a disease tend to participate in similar biological functions, guilt-by-association strategies have proved effective at identifying disease genes \cite{huang_systematic_2018}.

Network-based, biomarker discovery methods exploit the guilt-by-association strategy to identify disease genes on GWAS data \cite{azencott_network-guided_2016}. In essence, each SNP has a measure of association with the disease, given by the experiment, and functionally biological relationships, given by a network built on prior knowledge. Then, the problem becomes finding a functionally-related set of genes that is highly associated with the disease. Different solutions have been proposed to this problem, often stemming from divergent different mathematical frameworks and considerations of what the optimal solution looks like. Some methods strongly constrain the problem to certain kinds of subnetworks. Such is the extreme case of LEAN \cite{gwinner_network-based_2016}, which focuses on star subnetworks, i.e. instances were both a gene and its direct interactors are associated with the disease. Other algorithms, like dmGWAS \cite{jia_dmgwas:_2011} and heinz \cite{dittrich_identifying_2008}, focus on interconnected genes with high association with the disease. However, they differ in their tolerance to the inclusion of lowly associated nodes, and the possible number of disconnected subnetworks in the solution. Lastly, other methods also consider the topology of the network, favoring solutions that are densely interconnected; such is the case of HotNet2 \cite{leiserson_pan-cancer_2015}, SConES \cite{azencott_efficient_2013}, and SigMod \cite{liu_sigmod:_2017}.

In this work, we analyze the effectiveness of these six methods to discover new biomarkers on GWAS data. We focus on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study of familial breast cancer conducted in the French population. After a classical GWAS approach, we use these network-based methods to recover additional breast cancer biomarkers. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus network of predisposition to familial breast cancer.

@@

\subsection{High-score subnetwork search algorithms}
\subsubsection{SNP and gene association}
\label{methods:node_score}
To measure association between a genotype and the phenotype, we performed a per-SNP 1~d.f.~\(\chi\)\textsuperscript{2} allelic test using PLINK v1.90 \cite{chang_second-generation_2015}. Then, we used VEGAS2v2 to compute the gene-level association score \cite{mishra_vegas2:_2015} from the SNP P-values. In order to map SNPs to genes we used their overlap on the genome: all SNPs located within the boundaries of a gene, \textpm 50 kb, were mapped to that gene. To compute the gene association we used the 10\% of SNPs linked to the gene with lowest P-values. We used the 62\,193 genes described in GENCODE 31 \cite{frankish_gencode_2019}, although only 54\,612 could be mapped to at least one SNP. Out of those, we focused exclusively on the 32\,767 that had a gene symbol. Out of the SNPs 197\,083 remaining after quality control, 164\,037 were mapped to at least one of these genes.

@@

\begin{description}
\item[{heinz}] The goal of heinz is to identify the highest-scored connected subgraph on the network \cite{dittrich_identifying_2008}. The authors propose a transformation of the genes' P-value into a score that is negative under no association with the phenotype, and positive value when there is. This transformation is achieved by modelling the distribution of P-values by a beta-uniform model (BUM) parameterized by the desired FDR. Thus formulated, the problem is NP-complete. To solve it efficiently it is re-casted as the Prize-Collecting Steiner Tree Problem (PCST), which seeks to select the connected subnetwork S that maximizes the \emph{profit} p(S):

@@

\begin{description}
\item[{SigMod}] SigMod aims at identifying the most densely connected gene subnetwork that is most strongly associated to the phenotype \cite{liu_sigmod:_2017}. It addresses an optimization problem similar to that of SConES (Equation \ref{eq:scones}), but using the adjacency matrix rather than the Laplacian matrix (Section \ref{methods:notation}), to quantify solutions containing many edges.

@@

SigMod presents three important differences with SConES. First it is designed for gene-gene networks. Second, by replacing the adjacency by the Laplacian matrix, it favors subnetworks containing many edges. SConES, instead, penalizes connections between the selected and unselected nodes. Third, it returns a single connected subnetwork, which it achieves by exploring a grid of hyperparameters and processing their respective solutions. Specifically, for the range of \(\lambda\) = \(\lambda\)\textsubscript{min}, \dots{}, \(\lambda\)\textsubscript{max} for the same \(\eta\), it prioritizes the solution with the largest change in size from \(\lambda\)\textsubscript{n} to \(\lambda\)\textsubscript{n+1}. Such a large change implies that the network is strongly interconnected. This results in one candidate solution for each \(\eta\), which are processed by removing any node not connected to any other. A score is assigned to each candidate solution by summing their node scores and normalizing by size. The candidate solution with the highest standardized score is the chosen solution. SigMod is implemented in an R package \cite{sigmod}.
\end{description}

@@

\subsubsection{SNP networks}
\label{methods:snp_network}
SConES \cite{azencott_efficient_2013} is the only of the studied methods designed to handle SNP networks. As in gene networks, two SNPs are linked in a SNP network when there is evidence of shared functionality between two SNPs. The authors suggested three ways of building these networks: connecting the SNPs consecutive in the genomic sequence (``GS network''); interconnecting all the SNPs mapped to the same gene, on top of GS (``GM network''); and interconnecting all SNPs mapped to two genes for which a protein-protein interaction exists (``GI network''). We focused on the GI network, as it is the network that fits better the scope of this article. However, at different stages of this work we also used GS and GM. For the GM network, we used the mapping described in Section \ref{methods:node_score}. For the GI network, we used the PPI as described in Section \ref{methods:gene_network}. For all three networks the node score used is the association of the individual SNPs with the phenotype; specifically, we used the 1~d.f.~\(\chi\)\textsuperscript{2}.

\subsection{Evaluation of methods}
\subsubsection{Classification accuracy of selected biomarkers}
\label{methods:comparison}
A desirable solution is one that is sparse, while offering a good predictive power on unseen samples. We evaluated the predicting power of the SNPs selected by the different methods through the performance of an L1-penalized logistic regression trained exclusively on those SNPs to predict the outcome (case/control). The L1 penalty helps to account for LD to reduce the number of SNPs included in the model (size of the active set), while improving the generalization of the classifier. The value of the regularization parameter, which controls both the magnitude and the sparsity of the coefficients, was set by cross-validation. To that end, we used the different network-methods on a random subset of 80\% of the samples. On this same subset we trained our classifier exclusively on the SNPs selected by a particular method. When the method retrieved a list of genes (all of them except SConES), all the SNPs mapped to any of those genes were used. Then we evaluated performance of the classifier on the remaining 20\% of the dataset. We repeated this procedure 5 times to estimate the average and the deviation of the different performance measures. The performance measures we used were: size of the solution, size of the active set, specificity, and sensitivity. The size of the active set provides an estimate of a plausible, more sparse solution with a comparable predictive power to the original solution.

Additionally, for each of the methods, we evaluated their stability and their runtime. The stability of an algorithm is its sensitivity to small changes of the input, and is measured using the Pearson's correlation between different runs as suggested in \cite{nogueira_measuring_2016}. To obtain a baseline, we also performed the procedure using all the SNPs.

\subsubsection{Biological relevance of the genes}
\label{methods:bcac}
An alternative way to evaluate the results is comparing our results to an external dataset. For that purpose, we recovered a list of 153 genes associated to familial breast cancer from DisGeNET \cite{pinero_disgenet:_2017}. Across this article we refer to these genes as \emph{breast cancer susceptibility genes}.

Additionally, we used the summary statistics from the Breast Cancer Association Consortium (BCAC) \cite{michailidou_genome-wide_2015}. BCAC has conducted one of the largest efforts in GWAS, involving over 120\,000 women of European ancestry. As opposed to GENESIS, samples were not selected based on family history, and hence the study is enriched in sporadic breast cancers. Another difference is that BCAC is a relatively heterogeneous study on a pan-European sample, while GENESIS is a homogeneous dataset focused on the French population. Despite these differences, there should be shared genetic architecture. On top of that, that overlap should become larger at the gene level. For that purpose, we computed the gene association as in Section \ref{methods:node_score}. The iCOGS array was used for genotyping in BCAC \cite{sakoda_turning_2013}, the same array as for GENESIS \cite{sinilnikova_genesis:_2016}. Although imputed data are available, we used exclusively the SNPs available on GENESIS after quality control to make the results comparable.

@@

\subsection{A conventional GWAS confirms that FGFR2 is strongly associated with familial breast cancer}
\label{results:conventional}

We conducted association analyses in the GENESIS dataset at both the SNP and the gene levels (Section \ref{methods:node_score}). Two genomic regions have a P-value lower than the Bonferroni threshold in chromosomes 10 and 16 (Supplementary Figure \ref{sfig:snp_gene_manhattan}A). The former overlaps with gene \emph{FGFR2}; the latter with \emph{CASC16}, and it is located near the protein-coding gene \emph{TOX3}. Variants in both \emph{FGFR2} and \emph{TOX3} were related to breast cancer susceptibility in other cohorts negative for \emph{BRCA1/2} \cite{rinella_genetic_2013}. Only the peak in chromosome 10 replicated in the gene-level analysis, with \emph{FGFR2} just above the threshold of significance (Supplementary Figure \ref{sfig:snp_gene_manhattan}B). These results show an overlap in the genetic architecture of the disease between the studied French population sample and other populations, especially at the gene level. In addition, there are other SNPs whose P-values, although higher than the conventional threshold of significance, show a strong association with familial breast cancer. The most prominent of such regions are 3p24 and 8q24, both of which have been associated to breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. We applied an L1-penalized logistic regression (Section \ref{methods:comparison}), a machine learning algorithm that searches a small subset of SNPs which provide good classification accuracy. Unconstrained by considerations about statistical significance, it selected 100 SNPs, from all the aforementioned regions and others (Supplementary Figure \ref{sfig:snp_gene_manhattan}C). Yet, it is unclear why those SNPs were selected, as emphasized by the low association of some of them, which further complicates the biological interpretation. In this regard, and in opposition to what would be predicted by the omnigenic model, the genes to which this SNPs map to (Section \ref{required}) are not interconnected in the HINT protein-protein interaction network (Section \ref{required}). This motivates exploring network methods which, by trading statistical significance for biological relevance, search susceptibility subnetworks. In fact, such methods provided comparable classification accuracies to L1-penalized logistic regression (Figure \ref{fig:benchmark}D), while providing more interpretable solutions.

@@

We applied six network methods to the GENESIS dataset (Section \ref{methods:methods}). As none of the networks examined by LEAN was significant (adjusted P-value < 0.05), we obtained five solutions (Supplementary Figure \ref{fig:solution_overview}, Supplementary Files 1 and 2): one for each of the remaining four gene-based methods (Section \ref{methods:gene_network}), and one for SConES GI (Section \ref{methods:snp_network}). The solutions are very heterogeneous (Table \ref{tab:gene_solutions} and Supplementary table \ref{tab:snp_solutions}), making hard to draw joint conclusions. HotNet2 produced the largest solution subnetwork with 440 genes. SConES GI failed to recover genes in the PPIN, but it recovered one genomic region mapped to RNA gene \emph{RNU6-420P}. All solution subnetworks except LEAN's are, on average, more strongly associated to breast cancer than the whole PPIN (median P-values $\ll 0.46$), despite containing genes with higher P-values (Supplementary Figure \ref{fig:solution_overview}A). This exemplifies the trade-off between statistical significance and biological relevance. However, there are nuances between solutions: heinz strongly favored highly associated genes, while dmGWAS is less conservative (median gene P-values 0.0012 and 0.19, respectively); SConES tended to select whole LD-blocks; and HotNet2 and SigMod were less likely to select lowly associated genes.

>>>>

\documentclass[twocolumn, 11pt, draft]{article}
\usepackage{amsmath,amssymb}
\usepackage[format=hang,labelfont=bf,textfont=small,singlelinecheck=false,justification=raggedright,margin={20pt,20pt},figurename=Fig.]{caption}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage{threeparttable}
\usepackage[numbers]{natbib}
\newcommand{\mean}[1]{$\overline{\mbox{#1}}$}
\newcommand{\median}[1]{$\hat{\mbox{#1}}$}

\title{Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer}
\author{Héctor Climente-González$^{1,2,3}$, Christine Lonjou$^{1,2,3}$, Fabienne Lesueur$^{1,2,3}$, \\
GENESIS Study collaborators, Dominique Stoppa-Lyonnet$^{4,5,6}$, \\
Nadine Andrieu$^{1,2,3}$, Chloé-Agathe Azencott$^{3,1,2}$}
\date{$^{1}$Institut Curie, PSL Research University, F-75005 Paris, France;\\
  $^{2}$INSERM, U900, F-75005 Paris, France;\\
  $^{3}$MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France;\\
  $^{4}$Service de Génétique, Institut Curie, F-75005 Paris, France;\\
  $^{5}$INSERM, U830, F-75005 Paris, France;\\
  $^{6}$Université Paris Descartes.\\
}
\hypersetup{
  pdfauthor={Héctor Climente-González, Christine Lonjou, Fabienne Lesueur, GENESIS Study collaborators, Dominique Stoppa-Lyonnet, Nadine Andrieu, Chloé-Agathe Azencott},
  pdftitle={Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer},
  pdfkeywords={},
  pdfsubject={},
  pdflang={English}}
\begin{document}

@@

\begin{abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. It does so by jointly considering the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study the utility of six network methods to discover new biomarkers for breast cancer susceptibility by searching subnetworks with higher association scores to this phenotype. We interrogate a familial breast cancer genome-wide association study focused on \emph{BRCA1/2} negative French women. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. Interestingly, a combination of solution subnetworks provided a concise subnetwork of 93 genes, enriched in known breast cancer susceptibility genes (\emph{BABAM1}, \emph{BLM}, \emph{CASP8}, \emph{FGFR2}, and \emph{TOX3}, Fisher's exact test P-value = $7.8 \times 10^{-5}$ ) and occupy more central positions in the network than average. Additionally, it includes subnetworks of mechanisms related to cancer, like protein folding (\emph{HSPA1A}, \emph{HSPA1B}, and \emph{HSPA1L}) or mitochondrial ribosomes (\emph{MRPS30}, \emph{MRPS31}, \emph{MRPS18B}). We also observed a general dysregulation in the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. By trading statistical astringency for biological meaningfulness, most network methods get more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. 
\end{abstract}

@@

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, the experimental setting also presents intrinsic challenges. Some of them stem from the high-dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with large and moderate effects can be detected. And it is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically Bonferroni. However, Bonferroni is known to be overly conservative when the statistical tests are correlated, as is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. On top of that, recent large-sampled studies suggest that most of the genome contributes to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The recently proposed omnigenic model \cite{boyle_expanded_2017} offers an explanation: genes are very functionally inter-related and influence each other's behavior, which allows alterations in most genes to impact the subset of ``core'' genes directly involved in a disease's mechanism. Hence, a comprehensive statistical framework which includes the structure of biological data might address the aforementioned issues.

In this regard many authors turn to network biology to handle the complex interplay of biomolecules that lead to disease \cite{furlong_human_2013}. As its name suggests, network biology models biology as a network, where the biomolecules under study, often genes, are nodes, and between them are the edges that link them. These functional relationships come from evidence that the genes jointly contribute to a biological function; for instance, their expression is correlated, or their products establish a protein-protein interaction. Under this view, complex diseases are not the consequence of a single altered gene, but of the interaction of multiple interdependent molecules \cite{barabasi_network_2011}. In fact, an examination of biological networks shows that disease genes have differential properties \cite{barabasi_network_2011} \cite{pinero_uncovering_2016}. This is particularly true for cancer driver genes, which tend to be key players in connecting different, densely-connected communities of genes. Additionally, as genes that contribute to a disease tend to participate in similar biological functions, studying the neighborhood of known disease genes is effective at identifying new ones \cite{huang_systematic_2018}.

Network-based, biomarker discovery methods exploit the aforementioned strategy to identify disease genes on GWAS data \cite{azencott_network-guided_2016}. In essence, each SNP has a score of association with the disease, given by the experiment, and functionally biological relationships, given by a network built on prior knowledge. Then, the problem becomes finding a functionally-related set of highly-scoring genes. Different solutions have been proposed to this problem, often stemming from divergent different mathematical frameworks and considerations of what the optimal solution looks like. Some methods strongly constrain the problem to certain kinds of subnetworks. Such is the extreme case of LEAN \cite{gwinner_network-based_2016}, which focuses on star subnetworks, i.e. instances were both a gene and its direct interactors are associated with the disease. Other algorithms, like dmGWAS \cite{jia_dmgwas:_2011} and heinz \cite{dittrich_identifying_2008}, focus on interconnected genes with high association with the disease. However, they differ in their tolerance to the inclusion of low scoring nodes, and the possible number of disconnected subnetworks in the solution. Lastly, other methods also consider the topology of the network, favoring solutions that are densely interconnected; such is the case of HotNet2 \cite{leiserson_pan-cancer_2015}, SConES \cite{azencott_efficient_2013}, and SigMod \cite{liu_sigmod:_2017}.

In this work, we analyze the effectiveness of these six network methods for biomarker discovery on GWAS data. While all of them capture susceptibility mechanisms resembling that postulated by the omnigenic model, they do so in different ways, and provide a representative view of the field. We focus on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study of familial breast cancer conducted in the French population. After a classical GWAS approach, we use these network-based methods to recover additional breast cancer biomarkers. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus network of predisposition to familial breast cancer.

@@

\subsection{High-score subnetwork search algorithms}
\subsubsection{SNP and gene association}
\label{methods:node_score}
To measure association between a genotype and the phenotype, we performed a per-SNP 1~d.f.~\(\chi\)\textsuperscript{2} allelic test using PLINK v1.90 \cite{chang_second-generation_2015}. Then, we used VEGAS2v2 to compute the gene-level association score \cite{mishra_vegas2:_2015} from the SNP P-values. In order to map SNPs to genes we used their overlap on the genome: all SNPs located within the boundaries of a gene, \textpm 50 kb, were mapped to that gene. To compute the gene association score we used the 10\% of SNPs linked to the gene with lowest P-values. We used the 62\,193 genes described in GENCODE 31 \cite{frankish_gencode_2019}, although only 54\,612 could be mapped to at least one SNP. Out of those, we focused exclusively on the 32\,767 that had a gene symbol. Out of the SNPs 197\,083 remaining after quality control, 164\,037 were mapped to at least one of these genes.

@@

\begin{description}
\item[{heinz}] The goal of heinz is to identify the highest-scored connected subgraph on the network \cite{dittrich_identifying_2008}. The authors propose a transformation of the genes' P-value into a score that is negative under no association with the phenotype, and positive value when there is. This transformation is achieved by modelling the distribution of P-values by a beta-uniform model (BUM) parameterized by the desired FDR. Thus formulated, the problem is NP-complete, and hence solving it would require a prohibitely long computational time. To solve it efficiently it is re-casted as the Prize-Collecting Steiner Tree Problem (PCST), which seeks to select the connected subnetwork S that maximizes the \emph{profit} p(S):

@@

\begin{description}
\item[{SigMod}] SigMod aims at identifying the highest-scoring, most densely connected gene subnetwork \cite{liu_sigmod:_2017}. It addresses an optimization problem similar to that of SConES (Equation \ref{eq:scones}), but using the adjacency matrix rather than the Laplacian matrix (Section \ref{methods:notation}), to quantify solutions containing many edges.

@@

SigMod presents three important differences with SConES. First it is designed for gene-gene networks. Second, by replacing the adjacency by the Laplacian matrix, it favors subnetworks containing many edges. SConES, instead, penalizes connections between the selected and unselected nodes. Third, it returns a single connected subnetwork, which it achieves by exploring a grid of hyperparameters and processing their respective solutions. Specifically, for the range of \(\lambda\) = \(\lambda\)\textsubscript{min}, \dots{}, \(\lambda\)\textsubscript{max} for the same \(\eta\), it prioritizes the solution with the largest change in size from \(\lambda\)\textsubscript{n} to \(\lambda\)\textsubscript{n+1}. Such a large change implies that the network is densely interconnected. This results in one candidate solution for each \(\eta\), which are processed by removing any node not connected to any other. A score is assigned to each candidate solution by summing their node scores and normalizing by size. The candidate solution with the highest standardized score is the chosen solution. SigMod is implemented in an R package \cite{sigmod}.
\end{description}

@@

\subsubsection{SNP networks}
\label{methods:snp_network}
SConES \cite{azencott_efficient_2013} is the only of the studied methods designed to handle SNP networks. As in gene networks, two SNPs are linked in a SNP network when there is evidence of shared functionality between two SNPs. The authors suggested three ways of building these networks: connecting the SNPs consecutive in the genomic sequence (``GS network''); interconnecting all the SNPs mapped to the same gene, on top of GS (``GM network''); and interconnecting all SNPs mapped to two genes for which a protein-protein interaction exists (``GI network''). We focused on the GI network, as it is the network that fits better the scope of this article. However, at different stages of this work we also used GS and GM for comparison. For the GM network, we used the mapping described in Section \ref{methods:node_score}. For the GI network, we used the PPI as described in Section \ref{methods:gene_network}. For all three networks the node score used is an association score of the individual SNPs with the phenotype (1~d.f.~\(\chi\)\textsuperscript{2}).

\subsection{Evaluation of methods}

We evaluated multiple properties of the different method through a 5-fold subsampling setting. We applied each method to 5 random subsets of the original dataset containing 80\% of the samples (\emph{train set}). When pertinent, we evaluated the solution on the remaining 20\% (\emph{test set}). We used the 5 repetitions to estimate the average and the deviation of the different measures.

\subsubsection{Properties ot the solution}
\label{methods:comparison}

We compared the runtime, the number of selected genes/SNPs, and the stability (sensitivity of the result to small changes in the input) of the different network methods. The stability was measured using the Pearson's correlation between different runs as suggested \citet{nogueira_measuring_2016}.

\subsubsection{Classification accuracy of selected biomarkers}
\label{methods:classifier}

A desirable solution offers good predictive power on unseen \emph{test} samples. We evaluated the predicting power of the SNPs selected by the different methods through the performance of an L1-penalized logistic regression classifier trained exclusively on those SNPs to predict the outcome (case/control). The L1 penalty helps to account for LD by reducing the number of SNPs included in the model (size of the active set), while improving the generalization of the classifier. This penalty was set by cross-validation. We applied each network method to each \emph{train} set, and trained the classifier on the \emph{train} set as well using only on the selected SNPs. When the method retrieved a list of genes (all of them except SConES), we considered as selected all the SNPs mapped to any of those genes. Then we evaluated the sensitivity and the specificity on the \emph{test set}. To obtain a baseline, we also trained the classifier on all the SNPs. We also evaluated the size of the active set, an estimate of a plausible, more sparse solution with a comparable predictive power to the original solution.

\subsubsection{Comparison to state-of-the-art}
\label{methods:bcac}
An alternative way to evaluate the results is comparing our results to an external dataset. For that purpose, we recovered a list of 153 genes associated to familial breast cancer from DisGeNET \cite{pinero_disgenet:_2017}. Across this article we refer to these genes as \emph{breast cancer susceptibility genes}.

Additionally, we used the summary statistics from the Breast Cancer Association Consortium (BCAC), a meta-analysis of case-control studies conducted in multiple countries which included 228\,951 women of European ancestry from the general population \cite{Michailidou2017}. Hence, a high proportion of breast cancer cases investigated in BCAC are sporadic, while GENESIS is a homogeneous dataset focusing on the French high-risk population attending the family cancer clinics. Despite these differences, there should be shared genetic architecture. On top of that, that overlap should become larger at the gene level. For that purpose, we searched for associated genes as in Section \ref{methods:node_score}. After imputation, it consisted of 13\,250\,642 SNPs. We used all of them to measure gene-level association, but only the SNPs available on GENESIS after quality control for the SNP-level comparisons.

@@

\subsection{A conventional GWAS confirms that FGFR2 is associated with familial breast cancer}
\label{results:conventional}

We conducted association analyses in the GENESIS dataset at both the SNP and the gene levels (Section \ref{methods:node_score}). Two genomic regions have a P-value lower than the Bonferroni threshold in chromosomes 10 and 16 (Supplementary Figure \ref{sfig:snp_gene_manhattan}A). The former overlaps with gene \emph{FGFR2}; the latter with \emph{CASC16}, and it is located near the protein-coding gene \emph{TOX3}. Variants in both \emph{FGFR2} and \emph{TOX3} were related to breast cancer susceptibility in other case-control studies \cite{Michailidou2017}, \emph{BRCA1/2} carriers studies \cite{Mulligan2011}, and in hereditary breast and ovarian cancer families negative for \emph{BRCA1/2} \cite{rinella_genetic_2013}. Only the peak in chromosome 10 replicated in the gene-level analysis, with \emph{FGFR2} just above the threshold of significance (Supplementary Figure \ref{sfig:snp_gene_manhattan}B). These results show an overlap in the genetic architecture of the disease between the studied French population sample and other populations, especially at the gene level. In addition, there are compelling regions having low, albeit not genome-wide significant, P-values. The most prominent are 3p24 and 8q24, both of which have been associated to breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. We applied an L1-penalized logistic regression (Section \ref{methods:classifier}), a machine learning algorithm that searches a small subset of SNPs which provide good classification accuracy. Unconstrained by considerations about statistical significance, it selected 100 SNPs, from all the aforementioned regions and others (Supplementary Figure \ref{sfig:snp_gene_manhattan}C). Yet, it is unclear why those SNPs were selected, as emphasized by the high P-value of some of them, which further complicates the biological interpretation. In this regard, and in opposition to what would be predicted by the omnigenic model, the genes to which this SNPs map to (Section \ref{methods:node_score}) are not interconnected in the PPIN (Section \ref{methods:gene_network}). This motivates exploring network methods which, by trading statistical significance for biological relevance, search susceptibility subnetworks. In fact, such methods provided comparable classification accuracies to L1-penalized logistic regression (Figure \ref{fig:benchmark}D), while providing more interpretable solutions.

@@

We applied six network methods to the GENESIS dataset (Section \ref{methods:methods}). As none of the networks examined by LEAN was significant (adjusted P-value < 0.05), we obtained five solutions (Supplementary Figure \ref{fig:solution_overview}, Supplementary Files 1 and 2): one for each of the remaining four gene-based methods (Section \ref{methods:gene_network}), and one for SConES GI (Section \ref{methods:snp_network}). The solutions are very heterogeneous (Table \ref{tab:gene_solutions} and Supplementary table \ref{tab:snp_solutions}), making hard to draw joint conclusions. HotNet2 produced the largest solution subnetwork with 440 genes. SConES GI failed to recover genes in the PPIN, but it recovered one genomic region mapped to RNA gene \emph{RNU6-420P}. All solution subnetworks except LEAN's have, on average, lower association P-values than the whole PPIN (median P-value $\ll 0.46$), despite containing genes with higher P-values (Supplementary Figure \ref{fig:solution_overview}A). This exemplifies the trade-off between statistical significance and biological relevance. However, there are nuances between solutions: heinz strongly favored high scoring genes, while dmGWAS is less conservative (median gene P-values 0.0012 and 0.19, respectively); SConES tended to select whole LD-blocks; and HotNet2 and SigMod were less likely to select low scoring genes.
